Physics-informed Implicit Neural Representation for Potential Field Geophysics

### 2023-08-26 paper3.tex
Improve figures and minor text
As sent to supervisor for rough draft.

\subsection{Geophysical sampling and regularisation}
\label{sec:geo_airborne}
Airborne or ground-based potential field surveys sample a magnetic (or gravitational) scalar field on a three-dimensional drape surface.
Cultural obstructions or topography within the survey extent cause this drape surface to be irregular, and the sample points scattered.
The resolvable detail within the sampled field is a function of the sampling density, and the distance from the source.
% It is common to define a nominally regular drape surface with these parameters in mind, while respecting the cost of acquisition.
To enable digital processing and storage of the sample data following acquisition, the scattered points are regularised to a quantised grid.
The Nyquist wavelength inherent to the selected grid cell size during regularisation limits the high-resolution information representable by the grid.
Because grid regularisation is one of the earliest steps in interpreting geophysical surveys, this information fails to be utilised in any subsequent task.
% Subsequent processes such as frequency filtering or directional gradients are similary constrained by the parameters of the regularisation.
% Additionally, storing the grid product requires amount of memory, to represent the value with adequate precision at each cell in the resulting grid.

@@

As denoted by the \(L^2\) operator \(\lVert{}\cdot{}\rVert{}_{2}\) and weighting \(\beta{}\), this criterion additionally penalises predictions that are not close to zero.

@@

\subsection{Network architecture}
The Coordinate MLP architecture comprises a simple fully connected network with \(N\) hidden layers, each with \(m\) features.
The sinusoidal activation of \parencite{sitzmann2019siren} is used. % for a baseline, and comparison is made to Gaussian and wavelet activations.
High quality implicit represenation has been achieved in contemporary INR research with only several hidden layers, typically between 3 and 8, and \(N = 8\) is used here.
The number of hidden features is set \(m = 256\).
Figure \cref{fig:arch} shows the structure of the architecture.

\begin{figure}[hbt]
    \label{fig:arch}
    \caption[Network Architecture]{The Coordinate MLP architecture used for the implicit neural represenation task.}
\end{figure}

@@

Implicit functions are trained using coordinate-value pairs, and for low-resolution images the entire cell coordinate and cell value dataset fits within GPU memory for batch gradient descent.
The Wolfe Creek meteorite impact crater case study line data has \SI{76420}{} sample points, with the Geoscience Australia provided grid product having a resolution of 446 \(\times{}\) 445 cells.
The line coordinate dataset easily fits into available graphics memory, and can be trained using batch gradient descent.
%However, training on large scattered aeromagnetic surveys with more than one million samples, memory becomes a constraint and mini-batch training is required.
Training was performed on an Intel i9-10900KF workstation with an Nvidia RTX 3090 24 GB GPU\@.
A highly accurate represenation can be learned within 1000 epochs using the Adam optimiser, for a duration in the order of 60 seconds.
The OneCycle learning rate scheduler \parencite{smithSuperconvergenceVeryFast2018} is used to provide fine tuning at low learning rates toward the end of training.

As described in \cref{sec:pinn}, the objective function comprises \(L^2\) loss between the input and predicted field values.
It does not include the Laplace loss as described above. %, summed with weighted Laplace's equation loss, \cref{eqn:cri}.
\begin{equation}
    \label{eqn:cri}
    \mathcal{L}_{Total} = \lVert{}u_{input} - u_{pred}\rVert{}_{2} % + \beta{} \lVert{}\frac{\partial^{2}{u_{pred}}}{\partial{}\phi{}_{pred}^{2}}\rVert{}_{2}
\end{equation}

\subsection{Derivative calculation}
A naive approach to calculating gradients with the INR would be to evaluate the represenation \(u = \phi{}(x, y, z)\) at a set of regular coordinates \({(x, y, z)}_1 \dots {(x,y,z)}_n\), and use numerical differentiation to calculate derivatives.
However, using Automatic Differentiation within Pytorch, the implicit represenation can be differentiated with respect to its inputs to provide gradients prior to being constrained by choice of quantisation.
This is to directly sample \(\frac{\partial{u}}{\partial{\phi{}{(x,y,z)}}}\) at \({(x, y, z)}_1 \dots {(x,y,z)}_n\).
Higher order derivatives can be also calculated using AD in this manner.

@@

\begin{figure}[hbt]
    \includegraphics[width=\textwidth]{fig/p3/grid_comparison.pdf}
    \caption[Grid prediction]{Comparison between the ground-truth grid processed by Geoscience Australia (left) and the constant level grid queried from the INR at \SI{40}{\m} (right).}
    \label{fig:grid}
\end{figure}

@@

Constant level grids can be queried from the INR at any height.
\Cref{fig:grid35} shows the case study extent queried at \SI{35}{m}.
While comparable features can still be interpreted within this level grid, the represenation learnt here fails to represent physically realistic continuation of the field.

\begin{figure}[hbt]
    %TODO query at AWAGS height and compare GA AWAGS grid.
    % TODO check if provided grid is AWAGS or MICRO levelled - maybe grid data myself.
    \includegraphics[width=\textwidth]{fig/p3/grid_comparison_35m.pdf}
    \caption[Grid prediction]{A constant level grid queried from the INR at the AWAGS datum height of \SI{35}{m} (right).}
    \label{fig:grid35}
\end{figure}

@@

% \begin{figure}[hbt]
%     % \includegraphics[width=\textwidth]{fig/p3/grid_subpct.pdf}
%     \caption[Grid subsampled]{Performance of level grid prediction shown at \qty{30}{\percent} and \qty{10}{\percent} random sampling of training data.}
%     \label{fig:gridpct}
% \end{figure}

\subsection{Spatial derivatives and filters}
Using AD, derivatives can be calculated between the input spatial coordinates and the output potential field represenation.
While the horizontal derivatives correspond well to numerical derivatives calculated on the Wolfe Creek ground truth grid (\cref{fig:hori_grad}), the vertical derivative suffers poor performance (\cref{fig:vert_grad}).
This correlates with the poor continuation performance of model when predicting level grids several meters or more beyond the nominal sample height.
This is inrepreted as the model learning a poor represenation of the vertical component of the potential field.

\begin{figure}[hbt]
    % \includegraphics[width=\textwidth]{fig/p3/hori_grad.pdf}
    \caption[Horizontal derivatives]{AD calculated horizontal derivatives of the INR\@. Left: Northing derivative, right: easting derivative}
    \label{fig:hori_grad}
\end{figure}

\begin{figure}[hbt]
    % \includegraphics[width=\textwidth]{fig/p3/vert_grad.pdf}
    \caption[Vertical derivative]{AD calculated vertical derivative of the INR.}
    \label{fig:vert_grad}
\end{figure}

Using the implicit representation, a selection of common geophysical filters are calculated prior to quantisation.

\begin{enumerate}
    \item ASIG
    \item tilt?
    \item idk?
\end{enumerate}

\section{Discussion}
The implicit represenation has strong representive capacity for the horizontal components of the potential field, generating level grids which are accurate against the ground truth grid.
Level grids can be generated at any height, with best performance achieved near the realised nominal survey altitude.
Performance worsens within approximately one standard deviation from that height (\SI{35} to \SI{45}{\m}), with gradual failure of the model outisde of that height range.
% Even when using a small subset of the dataset, this performance is retained.
It is interpreted from the poor quality of the vertical gradients and out-of-range predictions that the model is relying on learning a function between horizontally coplanar points, as opposed to a realistic vertical continuation function.
While quantised grids can be queried at any cell size, short wavelength information beyond the Nyquist limit of the survey line sampling is not predicted.
% TODO The addition of a PINN term for objective function ...

The case study survey presented is very small compared to typical regional aeromagnetic surveys.
A demonstration of the method on a large aeromagnetic survey, requiring minibatch training, is included in \cref{app:menzies}.
Training of this representation follows the parameters outlined in \cref{sec:training}, which were optimised for the Wolfe Creek case study, with the addition of a shuffled minibatch size of \SI{1024000}.
Attempts to train the CMLP with small batch sizes did not converge.%, reflecting the non-convergance result found for training on very large subsampling factors of the line dataset.

Implicit neural representation is a rapidly evolving field of research, and draws on contemporary deep learning improvements within related fields.
In addition to improving the high-frequency learning capacity, physics informed loss functions promise higher quality vertical continuation performance. 
The current method could trivially include additional target dimensions, which would allow representation learning of vector fields \(\mathbb{R}^{3}\) in three dimensions \(\mathbb{R}^{3}\).
This has potential application in tensor fields and associated predictive tasks.
These remain for future work.

\section{Conclusions}
High-resolution implicit neural represenation of potential field geophysics surveys is achieved with the use of coordinate MLP neural networks.
Resolution in the represenation is limited by network capacity, rather than initial grid cell size selection during regularisation.
Gradients and spatial filters can be caluclated using the full resolution of the learnt represenation, and gridded \emph{a posteri} of gradient or filter calculation.

@@

>>>>

\subsection{Geophysical sampling and regularisation}
\label{sec:geo_airborne}
Airborne or ground-based potential field surveys sample a magnetic (or gravitational) scalar field on a three-dimensional drape surface.
Cultural obstructions or topography within the survey extent cause this drape surface to be irregular, and the sample points scattered.
The resolvable detail within the sampled field is a function of the sampling density, and the distance from the source.
% It is common to define a nominally regular drape surface with these parameters in mind, while respecting the cost of acquisition.
To enable digital processing and storage of the sample data following acquisition, the scattered points are regularised to a quantised grid.
The Nyquist wavelength inherent to the selected grid cell size during regularisation limits the high-resolution information representable by the grid.
Because grid regularisation is one of the earliest steps in interpreting geophysical surveys, this information fails to be utilised in most tasks.
% Subsequent processes such as frequency filtering or directional gradients are similary constrained by the parameters of the regularisation.
% Additionally, storing the grid product requires amount of memory, to represent the value with adequate precision at each cell in the resulting grid.

@@

As denoted by the \(L^2\) operator \(\lVert{}\cdot{}\rVert{}_{2}\) and weighting \(\beta{}\), this criterion penalises predictions that are not close to zero.

@@

\subsection{Network architecture}
The Coordinate MLP architecture comprises a simple fully connected network with \(N\) hidden layers, each with \(m\) features.
The sinusoidal activation of \parencite{sitzmann2019siren} is used. % for a baseline, and comparison is made to Gaussian and wavelet activations.
High quality implicit represenation has been achieved in contemporary INR research with only several hidden layers, typically between 3 and 8, and \(N = 8\) is used here.
The number of hidden features is set \(m = 256\).
% Figure \cref{fig:arch} shows the structure of the architecture.

% \begin{figure}[hbt]
%     \label{fig:arch}
%     \caption[Network Architecture]{The Coordinate MLP architecture used for the implicit neural represenation task.}
% \end{figure}

@@

Implicit functions are trained using coordinate-value pairs, and for low-resolution images the entire cell coordinate and cell value dataset fits within GPU memory for batch gradient descent.
The Wolfe Creek meteorite impact crater case study line data has \SI{76420}{} sample points, with the Geoscience Australia provided grid product having a resolution of 446 \(\times{}\) 445 cells.
The line coordinate dataset easily fits into available graphics memory, and can be trained using batch gradient descent.
%However, training on large scattered aeromagnetic surveys with more than one million samples, memory becomes a constraint and mini-batch training is required.
Training was performed on an Intel i9-10900KF workstation with an Nvidia RTX 3090 24 GB GPU\@.
A highly accurate represenation can be learned within 1000 epochs using the Adam optimiser, for a duration in the order of 60 seconds.
The OneCycle learning rate scheduler \parencite{smithSuperconvergenceVeryFast2018} with a maximum rate of \(3\times{}10^{-4}\) is used which provides fine tuning at low learning rates toward the end of training.

As described in \cref{sec:pinn}, the objective function comprises \(L^2\) loss between the input and predicted field values.
It does not use the Laplace loss discussed above. %, summed with weighted Laplace's equation loss, \cref{eqn:cri}.
\begin{equation}
    \label{eqn:cri}
    \mathcal{L}_{Total} = \lVert{}u_{input} - u_{pred}\rVert{}_{2} % + \beta{} \lVert{}\frac{\partial^{2}{u_{pred}}}{\partial{}\phi{}_{pred}^{2}}\rVert{}_{2}
\end{equation}

\subsection{Derivative calculation}
A naive approach to calculating gradients with the INR would be to evaluate the represenation \(u = \phi{}(x, y, z)\) at a set of regular coordinates \({(x, y, z)}_1 \dots {(x,y,z)}_n\), and use numerical differentiation to calculate derivatives.
However, using Automatic Differentiation within Pytorch, the implicit represenation can be differentiated with respect to its inputs to provide gradients prior to being queried with a quantised grid.
That is, directly sample \(\frac{\partial{u}}{\partial{\phi{}{(x,y,z)}}}\) at \({(x, y, z)}_1 \dots {(x,y,z)}_n\).
Higher order derivatives can be also calculated using AD in this manner.

@@

\begin{figure}[hbt]
    \centering{}
    \includegraphics[width=1.\linewidth]{fig/p3/grid_comparison.pdf}
    \caption[Grid prediction]{Comparison between the ground-truth grid processed by Geoscience Australia (left) and the constant level grid queried from the INR at \SI{40}{\m} (right).}
    \label{fig:grid}
\end{figure}

@@

Constant level grids can be queried from the INR at any height.
\Cref{fig:grid35} shows the case study extent queried at \SI{35}{\m}.
While comparable features can still be interpreted within this level grid, the represenation learnt here fails to represent physically realistic continuation of the field.
It is conjectured that implementing a PINN criterion such as Laplacian loss \cref{eqn:cri_laplace} would improve the performance these upward or downward extracted slices.

\begin{figure}[hbt]
    \centering{}
    % TODO query at AWAGS height and compare GA AWAGS grid.
    % TODO check if provided grid is AWAGS or MICRO levelled - maybe grid data myself.
    \includegraphics[width=1.\linewidth]{fig/p3/grid_comparison_35m.pdf}
    \caption[Grid prediction]{A constant level grid queried from the INR at the height of \SI{35}{\m} (right).}
    % AWAGS is 80 m, but we see nothing useful at 80m
    \label{fig:grid35}
\end{figure}

@@

% \begin{figure}[hbt]
%     % \includegraphics[]{fig/p3/grid_subpct.pdf}
%     \caption[Grid subsampled]{Performance of level grid prediction shown at \qty{30}{\percent} and \qty{10}{\percent} random sampling of training data.}
%     \label{fig:gridpct}
% \end{figure}

\subsection{Spatial derivatives and filters}
Using AD, derivatives can be calculated between the input spatial coordinates and the output potential field represenation.
While horizontal derivatives queried from the INR correspond well to numerical derivatives calculated on the 40 m level grid (\cref{fig:hori_grad}), the vertical derivative suffers poor performance (\cref{fig:vert_grad}).
This correlates with the poor continuation performance of model when predicting level grids several meters or more beyond the nominal sample height.
This is interpreted as the model learning a poor represenation of the vertical component of the potential field.

\begin{figure}[hbt]
    \centering
    \includegraphics[width=1.\linewidth]{fig/p3/dh_comparison.pdf}
    \caption[Horizontal derivatives]{AD (left column) and numerical (middle column) horizontal derivatives of the INR\@. Residuals are shown in the right column.}
    \label{fig:hori_grad}
\end{figure}

\begin{figure}[hbt]
    \centering
    \includegraphics[width=1.\linewidth]{fig/p3/dv_comparison.pdf}
    \caption[Vertical derivative]{AD (left) and numerical (right) vertical derivatives of the INR\@.}
    \label{fig:vert_grad}
\end{figure}

% Using the implicit representation, a selection of common geophysical filters are calculated prior to quantisation.

% \begin{enumerate}
%     \item ASIG
%     \item tilt?
%     \item idk?
% \end{enumerate}

\section{Discussion}
The implicit represenation has strong representative capacity for the horizontal components of the potential field, allowing extraction of level grids which are accurate against the ground truth grid.
Level grids can be generated at any height, with best performance achieved near the realised nominal survey altitude.
Performance worsens within approximately one standard deviation from that height (\SI{35} to \SI{45}{\m}), with gradual failure of the model outisde of that height range.
% Even when using a small subset of the dataset, this performance is retained.
It is interpreted from the poor quality of the vertical gradients and out-of-range predictions that the model is relying on learning a function between horizontally coplanar points, as opposed to a realistic vertical continuation function.
While quantised grids can be queried at any cell size, short wavelength information beyond the Nyquist limit of the survey line sampling is not predicted.
% TODO The addition of a PINN term for objective function ...

% The case study survey presented is very small compared to typical regional aeromagnetic surveys.
% A demonstration of the method on a large aeromagnetic survey, requiring minibatch training, is included in \cref{app:menzies}.
% Training of this representation follows the parameters outlined in \cref{sec:training}, which were optimised for the Wolfe Creek case study, with the addition of a shuffled minibatch size of \SI{1024000}\@.
% Attempts to train the CMLP with small batch sizes did not converge.%, reflecting the non-convergance result found for training on very large subsampling factors of the line dataset.

Implicit neural representation is a rapidly evolving field of research, and draws on contemporary deep learning improvements within related fields.
In addition to improving the high-frequency learning capacity, physics informed loss functions may allow higher quality vertical continuation performance. 
The current method could trivially include additional target dimensions, which would allow representation learning of vector fields \(\mathbb{R}^{3}\) in three dimensions \(\mathbb{R}^{3}\).
This has potential application in tensor fields and associated predictive tasks, and these remain for future work.

\section{Conclusions}
High-resolution implicit neural represenation of potential field geophysics surveys is achieved with the use of coordinate MLP neural networks.
Resolution in the represenation is limited by network capacity, rather than initial grid cell size selection during regularisation.
Level grids can be extracted from these represenations, emerging as a novel method for regularising scatter geophysical sample data.
Gradients and spatial filters can be caluclated using the full resolution of the learnt represenation, and gridded \emph{a posteri} of gradient or filter calculation.

@@

% \section{Appendix A}
% \label{app:menzies} %TODO format as "Appendix"
% The Menzies South ... Pxxxx survey contains xxx sample points across yyyy line kilometres, and was flown at a nominal terrain clearance of zz m.
% While a high-quality result is attainable using batch gradient descent on a subset of the survey points, minibatch gradient descent is demonstrated for the CMLP used to learn the full survey INR\@.

\begin{figure}[hbt]
    % \includegraphics[width=\textwidth]{fig/p3/menzies.pdf}
    \caption[Menzies INR]{Level grid represenation using minibatch gradient descent.}
    \label{fig:vert_grad}
\end{figure}
